Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations

### 2021-11-23 analysis/paper/paper.Rmd
minor intro/discussion edits

The Upper Salmon River major population group (MPG) supports eight independent, extant spring/summer Chinook Salmon *Oncorhynchus tshawytscha* populations including Salmon River (above Redfish Lake), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish), Pahsimeroi River, Lemhi River, and North Fork Salmon River [@NOAA2017]. At least five of these eight populations must meet criteria set forth by @McElhany2000 and @ICTRT2007 for the MPG to be considered viable and for the recovery of the Snake River Evolutionary Significant Unit (ESU). Populations within the ESU have substantial cultural value, support downriver mainstem Snake and Columbia River commercial and subsistence fisheries, and support local fisheries and economies in years with sufficient abundance. All populations within the Upper Salmon River MPG have become depleted in recent decades. Declines in survival of juvenile Chinook Salmon have been attributed to the removal of beavers from the landscape (fur trade), mining activities, river simplification, water withdrawals, logging activities, urbanization, avian predation, proliferation of non-native species (e.g., non-native coastal rainbow trout *O. mykiss irideus* and brook trout *Salvelinus fontinalis*), warming streams and rivers, and modifications to downriver migration corridors (e.g., from hydropower projects). Moreover, the abundance of returning adults are further impacted by ocean and downriver harvests, poor ocean conditions, and changes to the spawning migration corridor. Each of these factors have contributed, to varying and unknown extents, to reduced adult escapement, the primary metric used to assess population viability. In response to the decline in Chinook Salmon abundance from the myriad human activities and associated habitat degradation, action agencies have attempted to improve juvenile survival and adult spawning conditions by investing in the rehabilitation of tributary ecosystems.

@@

>>>>

The Upper Salmon River major population group (MPG) supports eight independent, extant spring/summer Chinook Salmon *Oncorhynchus tshawytscha* populations including Salmon River (above Redfish Lake), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish), Pahsimeroi River, Lemhi River, and North Fork Salmon River [@NOAA2017]. All populations within the Upper Salmon River MPG have become depleted in recent decades. Declines in survival of juvenile Chinook Salmon have been associated with hydrosystem passage (@Schaller2014), ocean conditions and regional climate (@Welch2021; @Gosselin2021), and freshwater habitat modification.

All populations within the Upper Salmon River MPG have become depleted in recent decades. Declines in survival of juvenile Chinook Salmon have been attributed to the removal of beavers from the landscape (fur trade), mining activities, river simplification, water withdrawals, logging activities, urbanization, avian predation, proliferation of non-native species (e.g., non-native coastal rainbow trout *O. mykiss irideus* and brook trout *Salvelinus fontinalis*), warming streams and rivers, and modifications to downriver migration corridors (e.g., from hydropower projects). Moreover, the abundance of returning adults are further impacted by ocean and downriver harvests, poor ocean conditions, and changes to the spawning migration corridor. Each of these factors have contributed, to varying and unknown extents, to reduced adult escapement, the primary metric used to assess population viability. In response to the decline in Chinook Salmon abundance from the myriad human activities and associated habitat degradation, action agencies have attempted to improve juvenile survival and adult spawning conditions by investing in the rehabilitation of tributary ecosystems.

@@

At least five of the eight populations in the Upper Salmon MPG must meet criteria set forth by @McElhany2000 and @ICTRT2007 for the MPG to be considered viable and for the recovery of the Snake River Evolutionary Significant Unit. Populations with the ESU have substantial cultural value, support downriver mainstem Snake and Columbia River commercial and subsistence fisheries, and support local fisheries and economies in years with sufficient abundance.
